\documentclass[11pt]{article}

\usepackage[hyphens]{url}
\usepackage[colorlinks]{hyperref}
\usepackage[margin=1in]{geometry}
\usepackage[charter]{mathdesign}

\begin{document}

\begin{center}
	\textbf{\Large DS-GA 1014: Optimization and computational linear algebra for data science\\ \medskip Fall 2020}
\end{center}

\begin{itemize}
	\item \textbf{Instructor:} Léo Miolane
	\begin{itemize}
		\item e-mail: \url{lm4271@nyu.edu}. 
		\item Office hours (on Zoom): Mondays 3-4pm or by appointment.
	\end{itemize}
	
	\item \textbf{Course information}
	\begin{itemize}
		\item Lecture time and location: Thursdays 3:30pm-5:10pm in Room 101 at 19 West 4th Street.
	\item Course website: \url{https://leomiolane.github.io/linalg-for-ds.html}
	\end{itemize}
	
	\item \textbf{Course outline}
	\begin{itemize}
		\item This course covers the basics of optimization and computational linear algebra used in Data Science. About 66\% of the lectures will be about linear algebra and ~33\% about convex optimization. The first 5 lectures will cover basic linear algebra. Then we will study applications: Markov chains and PageRank, PCA and dimensionality reduction, spectral clustering, linear regression. Lastly, we will go over convex functions, optimality conditions and gradient descent.
	\end{itemize}
	
	\item \textbf{Grade breakdown}
	\begin{itemize}
		\item Quizzes 5\%, on Gradescope, due weekly on Thursdays at $2$pm.
		\item Written homework to be (scanned and) submitted as a single PDF file on Gradescope 40\%. Homeworks are available on the course's website after the corresponding lecture and usually due on Sundays at 11:00 pm of the next week: this let you $\sim$ 10 days to complete it.
		\item Midterm examination: 20\%. 
		\item Final examination: 35\%. 
		\item All exams are open books/notes.
		\item No extra credit projects will be offered.
	\end{itemize}
	
	\item \textbf{Homework, quiz and examination policy}
	\begin{itemize}
		\item The lowest quiz score and lowest written homework scores will be dropped.
		\item Missed quizzes will not be made up.
		\item Late homework will not be accepted. This includes homework which the file metadata indicates was created before the deadline but which was not submitted on time. 
		\item Other homework extensions may be granted in cases when the lecturer has good reason to believe that a student may have difficulty completing homework on time even with proper time planning (e.g. joined the course late or prolonged illness). Please e-mail the lecturer well in advance and explain your reason(s) if you would like to request other homework extensions, and note that other homework extensions are not automatically granted.
		\item You are permitted to discuss homework with fellow students, but homework must be written in your own words. Please list on your submission the students you work with for the homework (this will not affect your grade).
		\item Homeworks contain extra-credit problems (indicated with a $\star$).They will not (directly) contribute to your score of this homework. However, for every $4$ extra credit questions successfully answered your lowest homework score get replaced by a perfect score.
		\item The Midterm and final exam will be take-home exams in limited time. This means that once you have access to the exam questions, you will have a limited time to upload your work.
	\end{itemize}
	

	
\item \textbf{Computing the course's grade}
	\begin{itemize}
	\item The grades of the different evaluations will be converted to a $0-100$ scale and averaged:
		$$
		{\rm Course \ grade} = 5\%\, ({\rm Quiz}) 
		+ 40\%\, ({\rm Homework})
		+ 20\%\, ({\rm Midterm})
		+ 35\%\, ({\rm Final})
		$$
	\item Your letter grade for the course (the one that will appear on Albert and your transcripts) is then obtained according to:
		\begin{center}
		\begin{tabular}{|c|c|}
			\hline
			Grade & Interval \\
			\hline
			A & $\geq 93$ \\
			A- & $90 \leq \cdot < 93$ \\
			B+ & $87 \leq \cdot < 90$ \\
			B & $83 \leq \cdot < 87$ \\
			B- & $80 \leq \cdot < 83$ \\
			C+ & $77 \leq \cdot < 80$ \\
			C & $73 \leq \cdot < 77$ \\
			C- & $70 \leq \cdot < 73$ \\
			D+ & $67 \leq \cdot < 70$ \\
			D & $63\leq \cdot < 67$ \\
			D- & $60 \leq \cdot < 63$ \\
			F & $< 60$\\
			\hline
		\end{tabular}
		\end{center}
	\item The grades of the quizzes/homeworks will not be curved. Only the grades of the midterm and final are going to be curved.
	\item The curve will works as follows. If the mean $M$ of the grades of the exam (midterm or final) is below $80$ out of $100$, then every student will receive extra $80-M$ points to its grade. In case the new grade exceeds $100$, it will be changed back to $100$.
\end{itemize}

	\item \textbf{In case the course switch to fully remote instruction}

		If for some reason the course has to switch to fully remote instructions, lectures and recitations will all be on Zoom according to the same schedule. Since quizzes, homeworks and exams are already 'remote', nothing will change on that side.

	\item \textbf{Academic integrity}
	\begin{itemize}
		\item You are expected to follow codes of academic integrity as specified by the university and the College of Arts and Sciences:
		\begin{itemize}
			\item \url{https://www.nyu.edu/about/policies-guidelines-compliance/policies-and-guidelines/academic-integrity-for-students-at-nyu.html}
			\item \url{https://cas.nyu.edu/content/nyu-as/cas/academic-integrity.html}
			\item \url{http://cas.nyu.edu/content/nyu-as/cas/academic-integrity/honor-code.html}
		\end{itemize}
	\end{itemize}
\end{itemize}

\end{document}
