\documentclass[11pt,nocut]{article}

\usepackage{../latex_style/packages}
\usepackage{../latex_style/notations}

\title{\vspace{-2.0cm}%
	Optimization and Computational Linear Algebra for Data Science\\
Final review problems}
%\author{Léo \textsc{Miolane} \ $\cdot$ \ \texttt{leo.miolane@gmail.com}}
\date{}
%\setcounter{section}{*}

\begin{document}
\maketitle
%\input{./preamble_homeworks.tex}
\begin{center}
	{\Large
		For review exercises on linear algebra, look at last year's final review exercises (available the course's website).
	}
\end{center}

\vspace{1mm}

\begin{problem}
	Let $A \in \R^{n \times m}$. Let $\sigma_1(A)$ be the largest singular value of $A$.
	Show that
	$$
	\sigma_1(A) = \max_{\|x\|=1} \|Ax\|.
	$$
\end{problem}

\vspace{2mm}

\begin{problem}
	Let $A \in \R^{n \times m}$. Show that $A^{\sT} A$  and $A A^{\sT}$ have the same non-zero eigenvalues.
\end{problem}

\vspace{2mm}

\begin{problem}[True or false?]
	For each of the following statement, say if they are true or false and justify your answer.
	\begin{itemize}
		\item For all $A \in \R^{n \times n}$, if $\lambda$ is an eigenvalue of $A$ then $\lambda^2$ is an eigenvalue of $A^2$.
		\item For all $A \in \R^{n \times n}$, if $\sigma$ is a singular value of $A$ then $\sigma^2$ is a singular value of $A^2$.
		\item For all symmetric matrix $A \in \R^{n \times n}$ the eigenvalues of $A$ are singular values of $A$.
	\end{itemize}
\end{problem}

\begin{problem}
	Let $A \in \R^{n \times m}$. Show that for all $u \in \Im(A)$ and for all $v \in \Ker(A^{\sT})$ we have
	$$
	\langle u,v \rangle = 0.
	$$
\end{problem}

%\begin{problem}
	%Recall that the infinity norm of $x \in \R^n$ is defined by $\|x\|_{\infty} = \max(|x_1|, \dots, |x_n|)$. Show that
	%$$
	%\|x\|_{\infty} \leq \|x\|_2 \leq \sqrt{n} \|x\|_{\infty}.
	%$$
%\end{problem}

%\begin{problem}
	%Let $A=U \Sigma V^{\sT}$ denote the singular value decomsition of $A \in \R^{n \times n}$. Using this decomposition, construct the matrix of the orthogonal projection onto $\Im(A)$.
%\end{problem}

\begin{problem}
	Let $A \in \R^{n \times m}$. Show that if $A$ has linearly independent columns, then $A^{\dagger} = (A^{\sT}A)^{-1} A^{\sT}$.
\end{problem}

\begin{problem}
	Which of the following functions $f: \R^n \to \R^n$ are convex? Justify your answer
	\begin{itemize}
		\item $f(x) = \|x\|^2$.
		\item $f(x) = Ax$, for some $A \in \R^{n \times n}$.
		\item $f(x) = \sum_{i=1}^n x_i^3$.
	\end{itemize}
\end{problem}

\begin{problem}
	Which of the following subset $S$ of $\R^n$ are convex? Justify your answer
	\begin{itemize}
		\item $S = \{x \in \R^n \, | \, \|x\|_{\infty} \leq 1\}$.
		\item $S = \{x \in \R^n \, | \, \|x\|_{1} \geq 1\}$.
		\item $S = \{x \in \R^n \, | \, \|Ax\| < 1\}$, for some $A \in \R^{n \times n}$.
	\end{itemize}
\end{problem}


\begin{problem}
	Show that we are performing PCA on $n$ data points $a_1, \dots, a_n \in \R^d$ and keep only the first $k < d$ principal components of each point. We store the dimensionally reduced dataset in a $n \times k$ matrix $B$, where $B_{i,j}$ is the $j^{\rm th}$ principal component of the point $a_i$. Show that the columns of $B$ are orthogonal.
\end{problem}

\newpage
\begin{problem}[True of false?]
	For each of the following statement, say if they are true or false and justify your answer.
	\begin{itemize}
		\item If a continuous function $f:\R \to \R$ has a unique minimizer then $f$ is convex.
		\item If a continuous function $f:\R \to \R$ is such that there exists $x_0$ such that $f$ is decreasing on $(-\infty,x_0]$ and increasing on $[x_0, + \infty)$ then $f$ is convex.
		\item A twice differentiable function $f: \R \to \R$ whose derivative $f'$ is non-decreasing is convex.
	\end{itemize}
\end{problem}


\begin{problem}
	Let $f: \R^n \to \R$ be a convex, differentiable function. Assume that there exist $x,y \in \R^n$ such that $\nabla f(x) = \nabla f(y) = 0$. Show that $\nabla f(\frac{1}{2}(x+y)) = 0$.
\end{problem}


\begin{problem}
	Assume that we are doing linear regression with the least-squares cost
	$$
	f(x) = \|Ax - y\|^2
	$$
	where $A \in \R^{n \times d}$ and $y \in \R^n$. Should you normalize the dataset $A$ (that is, should we divide each column of $A$ by its norm) to get better results (smaller training error or smaller test error on new data points)?
	\\

	Suppose that we now want to use the lasso and minimize
	$$
	f(x) = \|Ax - y\|^2 + \lambda \|x\|_1
	$$
	for some $\lambda > 0$. Is there any reason why you might want to normalize the dataset in that case?
\end{problem}


\begin{problem}
	Compute the critical points of the following function and say if they are local minimizers, local maximizers or saddle points.
	$$
	f(x,y,z) = x^2 + y^2 - z^2
	\quad \text{and} \quad g(x,y) = 3x^2 + y^2 - 6x -4y - 10.
	$$
\end{problem}
\begin{problem}
	Solve the following constrained minimization problem (find all the solutions to these problems).
	\begin{enumerate}
		\item Minimize $x + y + z$ subject to $e^{-x} + e^{-y} + e^{-z} = 1$.
		\item Minimize $x^2 + y^2 + z^2$ subject to $xyz=1$.
	\end{enumerate}
\end{problem}

\begin{problem}
	Assume that we are doing standard gradient descent to minimize the least-square cost
	$$
	f(x) = \|Ax-y\|^2.
	$$
	Assume that the columns of  $A$ are linearly dependent, meaning that $\Ker(A) \neq \{0\}$. At which speed should gradient descent converge to the minimum?
	If now $\Ker(A) = \{0\}$, at which speed should gradient descent converge?
	By speed, we only ask about the dependence in $t$, the number of iterations, of the gap 
	$f(x_t) - \min f$, where $x_t$ is the position of gradient descent after $t$ iterations.
\end{problem}

\begin{problem}
	Let $A \in \R^{n \times d}$. Assume that the columns of $A$ are linearly independent. How many steps of Newton's method do you need to minimize
	$$
	\|Ax - y\|^2 \ ?
	$$
	($y \in \R^n$ is a fixed vector). Justify your answer.
\end{problem}


\begin{problem}
	When running stochastic gradient descent, what are upsides and downsides of having a rapidly decaying learning rate?
\end{problem}

\vspace{1cm}
\centerline{\pgfornament[width=7cm]{87}}

%\bibliographystyle{plain}
%\bibliography{./references.bib}
\end{document}
