\documentclass[11pt,nocut]{article}

\usepackage{../../latex_style/packages}
\usepackage{../../latex_style/notations}

\title{\vspace{-2.0cm}%
	Optimization and Computational Linear Algebra for Data Science\\
Homework 9: Convex functions}
%\author{Léo \textsc{Miolane} \ $\cdot$ \ \texttt{leo.miolane@gmail.com}}
\date{\vspace{-1cm}Due on November 15, 2020}
\setcounter{section}{9}

\begin{document}
\maketitle
\input{./preamble_homeworks.tex}

%DRAFT

%\color{green}
\vspace{4mm}

\begin{problem}[2 points]
	Let $f: \R^n \to \R$ be a convex function. We assume that the minimum $m \defeq \min_{x \in \R^n} f(x)$ of $f$ on $\R^n$ is finite, and that the set of minimizers of $f$
$$
\mathcal{M} \defeq \big\{ v \in \R^n \, \big| \, f(v) = m \big\}
$$
is non-empty.

	\begin{enumerate}[label=\normalfont(\textbf{\alph*})]
		\item Show that $\mathcal{M}$ is a convex set.
		\item Show that if $f$ is strictly convex, then $\mathcal{M}$ has only one element.
	\end{enumerate}

\end{problem}

\vspace{4mm}

\begin{problem}[2 points]
	Let $M \in \R^{n \times n}$ be a symmetric matrix, $b \in \R^n$ and $c \in \R$.
	For $x \in \R^n$ we define
	$$
	f(x) = x^{\sT} M x + \langle x,b \rangle + c.
	$$
	$f$ is called a quadratic function.
	\begin{enumerate}[label=\normalfont(\textbf{\alph*})]
		\item Compute the gradient $\nabla f(x)$ and the Hessian $H_f(x)$ at all $x \in \R^n$. Show that $f$ is convex if and only if $M$ is positive semi-definite.
		\item In this question, we assume $M$ to be positive semi-definite. Show that $f$ admits a minimizer if and only if $b \in \Im(M)$.
	\end{enumerate}
\end{problem}

\newpage

\begin{problem}[3 points]\label{prob:strongly_convex}
	We say that a function $f: \R^n \to \R$ is strongly convex if there exists $\alpha >0$ such that the function $x \mapsto f(x) - \frac{\alpha}{2} \|x\|^2$ is convex. In other words, $f$ is strongly convex if there exists $\alpha > 0$ and a convex function $g: \R^n \to \R$ such that 
	$$
	f(x) = g(x) + \frac{\alpha}{2} \|x\|^2.
	$$
	
	\begin{enumerate}[label=\normalfont(\textbf{\alph*})]
		\item Show that a strongly convex function is strictly convex. (Hint: start by showing that $x \mapsto \|x\|^2$ is strictly convex).
		\item Let $\varphi:\R^n \to \R$ be a twice differentiable function. Show that $\varphi$ is strongly convex if and only if there exists $\alpha >0$ such that for all $x \in \R^n$ the eigenvalues of $H_{\varphi}(x)$ are greater or equal than $\alpha$.
	\end{enumerate}
\end{problem}


\vspace{4mm}

\begin{problem}[3 points]
	Let $A \in \R^{n \times m}$ and $y \in \R^n$.
	For $x \in \R^m$ we define
	$$
	f(x) = \|Ax-y\|^2.
	$$
	\begin{enumerate}[label=\normalfont(\textbf{\alph*})]
		\item Compute the gradient $\nabla f(x)$ and the Hessian $H_f(x)$ at all $x \in \R^m$. Show that $f$ is convex.
		\item Show that if $\rank(A)<m$, then $f$ is not strictly convex.
		\item Show that is $\rank(A)=m$, then $f$ is strongly convex (use the definition and results of Problem \ref{prob:strongly_convex}).
	\end{enumerate}
\end{problem}

\vspace{4mm}

\begin{problem}[$\star$]
	\textbf{Notation:} For a symmetric matrix $M \in \R^{n \times n}$ we denote respectively $\lambda_{\rm min}(M)$ and $\lambda_{\rm max}(M)$ the smallest and largest eigenvalue of $M$.
	\\

	Let $f: \R^n \to \R$ be a function that is twice continuously differentiable. We assume that 
	$$
	\gamma \defeq \inf_{x \in \R^n} \lambda_{\rm min}(H_f(x))
	\qquad \text{and} \qquad
	L \defeq \sup_{x \in \R^n} \lambda_{\rm max}(H_f(x))
	$$
	are both finite. Show that for all $x,h \in \R^n$:
	$$
	f(x) + \langle \nabla f(x), h \rangle + \frac{\gamma}{2} \|h\|^2
	\leq f(x+h)
	\leq
	f(x) + \langle \nabla f(x), h \rangle + \frac{L}{2} \|h\|^2.
	$$

\end{problem}

\vspace{1cm}
\centerline{\pgfornament[width=7cm]{87}}

%\bibliographystyle{plain}
%\bibliography{./references.bib}
\end{document}
