\documentclass[11pt,nocut]{article}

\usepackage{../../latex_style/packages}
\usepackage{../../latex_style/notations}

\title{\vspace{-2.0cm}%
	Optimization and Computational Linear Algebra for Data Science\\
Homework 7: Principal component analysis}
%\author{Léo \textsc{Miolane} \ $\cdot$ \ \texttt{leo.miolane@gmail.com}}
\date{\vspace{-1cm}Due on October 25, 2020}
\setcounter{section}{7}

\begin{document}
\maketitle
\input{./preamble_homeworks.tex}

%DRAFT
\vspace{5mm}

%\color{green}
\begin{problem}[2 points]
	We say that a symmetric matrix $M \in \R^{n \times n}$ is positive definite if for all \textbf{non-zero} $x \in \R^n$,
	$$
	x^{\sT} M x > 0.
	$$
	If a matrix $M$ is positive definite, then $M$ is also positive semi-definite, but the converse is not true. One of the goals of this problem is to prove one of the implications of Proposition~1.2 of the notes (Lecture~7). You are of course not allowed to use this proposition to solve this problem.
	\begin{enumerate}[label=\normalfont(\textbf{\alph*})]
		\item Let $M \in \R^{n \times n}$ be a positive definite matrix. Show that its eigenvalues are all strictly positive and that $M$ is invertible. 
		\item Let $M \in \R^{n \times n}$ be a symmetric matrix. Show that there exists $\alpha > 0$ such that the matrix $M + \alpha \Id_n$ is positive definite.
	\end{enumerate}
\end{problem}

\vspace{5mm}

\begin{problem}[3 points]
	Using PCA, we reduce the dimension of a dataset $a_1, \dots, a_n \in \R^d$ of mean zero, to get a <<dimensionally reduced dataset>> $b_1 , \dots, b_n \in \R^k$, for some $1 \leq k \leq d$.
	\begin{enumerate}[label=\normalfont(\textbf{\alph*})]
		\item Show that the dataset $b_1, \dots, b_n$ is centered: $\sum_{i = 1}^n b_i = 0$.
		\item Show that for all $i,j \in \{1, \dots, n\}$, we have
			$$
				\|b_i - b_j\| \leq \|a_i - a_j\| \,.
			$$
			This means that PCA shrinks the distances.
		\item For $i \in \{1, \dots, k\}$ we let
			$$
			f^{(i)} = (b_{1,i}, b_{2,i}, \dots, b_{n,i}) \in \R^n
			$$
			be the vector made of all $i^{\rm th}$ components of the vectors $b_1, \dots, b_n$.
			Show that for $i \neq j$, $f^{(i)} \perp f^{(j)}$. This means that the new features computed using PCA are uncorrelated.
	\end{enumerate}
\end{problem}

\newpage

\begin{problem}[2 points]
	Let $A \in \R^{n \times m}$. The Singular Values Decomposition (SVD) tells us that there exists two orthogonal matrices $U \in \R^{n \times n}$ and $V \in \R^{m \times m}$ and a matrix $\Sigma \in \R^{n \times m}$ such that $\Sigma_{1,1} \geq \Sigma_{2,2}  \geq \cdots \geq 0$ and $\Sigma_{i,j} = 0$ for $i\neq j$
	$$
	A = U \Sigma V^{\sT}.
	$$
	The columns $u_1, \dots, u_n$ of $U$ (respectively the columns $v_1, \dots, v_m$ of $V$) are called the left (resp.\ right) singular vectors of $A$. The non-negative numbers $\sigma_i \defeq \Sigma_{i,i}$ are the singular values of $A$. Moreover we also know that $r \defeq \rank(A) = \# \{i \, | \, \Sigma_{i,i} \neq 0 \}$.
	\begin{enumerate}[label=\normalfont(\textbf{\alph*})]
		\item Let 
			$\widetilde{U} = \!
			{\small \begin{pmatrix}
					| & & | \\
					u_1 & \!\cdots\! & u_r \\
					| & & | 
			\end{pmatrix} } \! \in \! \R^{n \times r}$ ,
			$\widetilde{V} = \!
			{\small \begin{pmatrix}
					| & & | \\
					v_1 & \!\cdots\! & v_r \\
					| & & | 
			\end{pmatrix} } \! \in \! \R^{m \times r}$ and
			$\widetilde{\Sigma} = \Diag(\sigma_1, \dots, \sigma_r) \in \R^{r \times r}$.
			Show that $A = \widetilde{U}\widetilde{\Sigma}\widetilde{V}^{\sT}$.
		\item Give orthonormal bases of $\Ker(A)$ and $\Im(A)$ in terms of the singular vectors $u_1, \dots, u_n, v_1, \dots , v_m$.
	\end{enumerate}
\end{problem}

\vspace{5mm}

\begin{problem}[3 points]
	You have been given a mysterious dataset that may contain important informations! This dataset is a collection of $n=6344$ points of dimension $d=1000$.
	Investigate the structure of this dataset using PCA/plots... , and find out if the dataset contains any information.
	\\

	The \texttt{zip} file \texttt{mysterious\_data.zip} contains a text file containing the $6344\times 1000$ data matrix.
	The \emph{Jupyter notebook} \texttt{mysterious\_data.ipynb} contains a function to read the text file.
\\

	You are not allowed to use any builtin PCA function: you have to do the all process by yourself (centering the data, computing the covariance matrix...). Of course, for computing eigenvalues/eigenvectors you will need to use the numpy library.
	The numpy function \texttt{numpy.linalg.eigh} is great to compute eigenvalues and eigenvectors of a symmetric matrix. 
	\\

	\textbf{It is intended that you code in Python and use the provided Jupyter Notebook. Please only submit a pdf version of your notebook (right-click $\to$ `print' $\to$ `Save as pdf').}
\end{problem}


\vspace{5mm}

\begin{problem}[$\star$]
	Let $M \in \R^{n \times n}$ be a symmetric matrix. Let $\lambda_1 \geq \dots \geq \lambda_n$ be the eigenvalues of $M$. Show that for all $d \leq n$:
	$$
	\max_{\substack{U \in \R^{n\times d} \\ U^{\sT} U = \Id_d}} \Tr(U^{\sT} M U) = \sum_{i=1}^d \lambda_i.
	$$
\end{problem}
\vspace{1cm}
\centerline{\pgfornament[width=7cm]{87}}

%\bibliographystyle{plain}
%\bibliography{./references.bib}
\end{document}
