\documentclass[11pt,nocut]{article}

\usepackage{../latex_style/packages}
\usepackage{../latex_style/notations}

\title{\vspace{-2.0cm}%
	Optimization and Computational Linear Algebra for Data Science\\
	Lecture 2: Linear transformations}
	\author{Léo \textsc{Miolane} \ $\cdot$ \ \texttt{leo.miolane@gmail.com}}
\date{\today}

\begin{document}
\maketitle
\input{../preamble.tex}
\section{General definitions}

We present below the abstract mathematical definition of a vector space. 
\textbf{Please do not try to memorize it!} 
Simply remember that a vector space is a set whose elements are called \emph{vectors}, that one can add vectors together and multiply them by real numbers called \emph{scalars}.
\begin{definition}[Vector space]
	A vector space (over $\R$) consists of of a set $V$ (whose elements are called vectors) and two operations $+$ and $\cdot$ that verify:
	\begin{enumerate}
		\item The sum of two vectors is a vector: for all $\vec{x},\vec{y} \in V$ we have $\vec{x}+\vec{y} \in V$.
		\item The vector sum is commutative and associative. For all $\vec{x},\vec{y},\vec{z} \in V$ we have
			$$
			\vec{x}+\vec{y} = \vec{y} + \vec{x} \qquad \text{and} \qquad \vec{x} + (\vec{y} + \vec{z}) = (\vec{x}+\vec{y}) + \vec{z}.
			$$
		\item There exists a zero vector $\vec{0} \in V$ that verifies $\vec{x} + \vec{0} = \vec{x}$ for all $\vec{x} \in V$.
		\item For all $\vec{x} \in V$, there exists $\vec{y} \in V$ such that $\vec{x} + \vec{y} = \vec{0}$. Such $\vec{y}$ is called the additive inverse of $\vec{x}$ and is written $- \vec{x}$.
		\item Scalar multiplication: for all $\vec{x} \in V$ and all $\alpha \in \R$, $\alpha \cdot \vec{x} \in V$.
		\item Identity element for scalar multiplication: $1 \cdot \vec{x} = \vec{x}$ for all $\vec{x} \in V$.
		\item Compatibility between scalar multiplication and the usual multiplication: for all $\alpha,\beta \in \R$ and all $\vec{x} \in V$, we have
			$$
			\alpha \cdot(\beta \cdot \vec{x}) = (\alpha \beta) \cdot \vec{x}.
			$$
		\item Distributivity: for all $\alpha,\beta \in \R$ and all $\vec{x},\vec{y} \in V$,
			$$
			(\alpha + \beta) \cdot \vec{x} = \alpha \cdot \vec{x} + \beta \cdot \vec{y}
			\qquad \text{and} \qquad
			\alpha \cdot (\vec{x} + \vec{y}) = \alpha \cdot \vec{x} + \alpha \cdot \vec{y}.
			$$
	\end{enumerate}
\end{definition}

From now we will ignore $\cdot$ and simply write $\alpha \vec{x}$ instead of $\alpha \cdot \vec{x}$.

\begin{example}
	\leavevmode
	\begin{itemize}
		\item The set $V = \R^n$ endowed with the usual vector addition $+$
			$$
			(x_1, \dots, x_n) + (y_1, \dots, y_n) = (x_1 + y_1, \dots, x_n + y_n)
			$$
			and the usual scalar multiplication $\cdot$
			$$
			\alpha \cdot (x_1, \dots, x_n)= (\alpha x_1, \dots, \alpha x_n)
			$$
			is a vector space.
		\item The set $V = \cF(\R,\R) \defeq \{ f \, | \, f: \R \to \R \}$ of all functions from $\R$ to itself endowed with the addition $+$ and the scalar multiplication $\cdot$ defined by
			$$
			\begin{array}{rlll}
			f+g :& \R & \to & \R \\
				 &t & \mapsto & f(t) + g(t)
			\end{array}
			\qquad
			\text{and}
			\qquad
			\begin{array}{rlll}
			\alpha \cdot f :& \R & \to & \R \\
							&t & \mapsto & \alpha f(t)
			\end{array}
			$$
			is a vector space.
	\end{itemize}
\end{example}

\begin{definition}[Subspace]
	We say that a non-empty subset $S$ of a vector space $V$ is a \emph{subspace} is it is stable by addition and multiplication by a scalar, that is if
	\begin{enumerate}[label=(\roman*),noitemsep]
		\item for all $x,y \in S$ we have $x + y \in S$,
		\item for all $x \in S$ and all $\alpha \in \R$ we have $\alpha x \in S$.
	\end{enumerate}
\end{definition}

Notice that a subspace is also a vector space!

\section{Linear dependency}

\begin{definition}[Span]
	Given vectors $x_1, \dots x_k \in V$ the \emph{span} of $x_1, \dots, x_k$ is the set of vectors that can be written as a linear combination of these:
	$$
	\Span(x_1, \dots, x_n) = 
	\Big\{ \alpha_1 x_1 + \cdots + \alpha_k x_k \, \Big| \, \alpha_1, \dots, \alpha_k \in \R \Big\}.
	$$
\end{definition}

One can easily verify (exercise!) that $\Span(x_1, \dots, x_n)$ is a subspace of $V$.

\begin{definition}[Linear dependency]
	Vectors $x_1, \dots x_k \in V$ are \emph{linearly dependent} is there exists $\alpha_1, \dots, \alpha_k \in \R$ \textbf{that are not all zero} such that 
	$$
	\alpha_1 x_1 + \cdots + \alpha_k x_k = 0.
	$$
	They are said to be \emph{linearly independent} otherwise.
\end{definition}

Saying that $x_1, \dots, x_k$ are linearly dependent precisely means that one of the vectors $x_1, \dots, x_k$ can be obtained as a linear combination of the others. 
Indeed if $x_1, \dots, x_k$ are linearly dependent, then we can find $\alpha_1, \dots, \alpha_k \in \R$ that are not all zero (there exists $i$ such that $\alpha_i \neq 0$) such that $\alpha_1 x_1 + \cdots + \alpha_k x_k = 0$. This leads to
$$
x_i = \sum_{j \neq i} \frac{- \alpha_j}{\alpha_i} x_j,
$$
i.e.\ the vector $x_i$ can be expressed as a linear combinations of the vectors $x_j$ for $j \neq i$.
Conversely if we have for some $i$, and $\alpha_1, \dots, \alpha_k \in \R$
$$
x_i = \alpha_1 x_1 + \cdots + \alpha_{i-1} x_{i-1} + \alpha_{i+1} x_{i+1} + \cdots \alpha_k x_k = 0.
$$
then $\alpha_1 x_1 + \cdots + \alpha_{i-1} x_{i-1} - x_i + \alpha_{i+1} x_{i+1} + \cdots \alpha_k x_k = 0$ which gives that $x_1, \dots, x_k$ are linearly dependent. 
\\

\begin{theorem}\label{th:lin_dep}
	Let $v_1, \dots, v_n \in V$ and suppose that we have vectors $x_1, \dots, x_k \in V$ such that $k > n$ and $x_i \in \Span(v_1, \dots, v_n)$ for all $i \in \{1, \dots, k\}$.
	Then $x_1, \dots, x_k$ are linearly dependent.
\end{theorem}
Theorem \ref{th:lin_dep} will be proved in Section \ref{sec:proof_lin_dep}.
\\

\begin{definition}[Basis]
	A family $(x_1, \dots, x_n)$ of vectors of $V$ is a basis of $V$ if
	\begin{enumerate}[label=(\roman*)]
		\item $x_1, \dots, x_n$ are linearly independent,
		\item $\Span(x_1, \dots, x_n) = V$.
	\end{enumerate}
\end{definition}

\begin{definition}[Dimension]
	If the vector space $V$ admits a base $(v_1, \dots, v_n)$, then any base of $V$ has also $n$ vectors. We say that $V$ has dimension $n$ and write $\dim(V) = n$.
\end{definition}
\begin{proof}
	We proceed by contradiction and assume that there exists two basis $(v_1, \dots, v_n)$ and $(x_1, \dots, x_k)$ of $V$ such that $k \neq n$. Without loss of generality we can assume that $k > n$. For $i = 1, \dots, k$ we have
	$$
	x_i \in V = \Span(v_1, \dots, v_n),
	$$
	because $(v_1, \dots, v_n)$ is a basis of $V$. We can therefore apply Theorem \ref{th:lin_dep} to get that $x_1, \dots, x_{n+1}$ are linearly dependent. This contradicts the fact that $(x_1, \dots, x_{k})$ is a basis.
\end{proof}
\\

\begin{proposition}[Coordinates]
	Let $(v_1, \dots, v_n)$ be a basis of $V$. Then for every $x \in V$ there exists a unique vector $(\alpha_1, \dots, \alpha_n) \in \R^n$ such that
	$$
	x = \alpha_1 v_1 + \dots + \alpha_n v_n.
	$$
	We say that $(\alpha_1, \dots, \alpha_n)$ are the coordinates of $x$ in the basis $(v_1, \dots, v_n)$.
\end{proposition}
\begin{proof}
	\textbf{Existence}. $(v_1, \dots, v_n)$ forms a basis of $V$ therefore  $V = \Span(v_1, \dots, v_n)$. We get that $x \in \Span(v_1, \dots, v_n)$ which gives that there exists $\alpha_1, \dots, \alpha_n$ such that $x = \alpha_1 v_1 + \cdot + \alpha_n v_n$.
	\\
	\textbf{Unicity}. Let $\alpha_1, \dots, \alpha_n, \beta_1, \dots, \beta_n \in \R$ such that
	\begin{align*}
		x &= \alpha_1 v_1 + \cdots + \alpha_n v_n = \beta_1 v_1 + \cdots + \beta_n v_n.
	\end{align*}
	This leads to
	$$
	(\alpha_1 - \beta_1) v_1 + \cdots + (\alpha_n - \beta_n) v_n = 0.
	$$
	The vectors $v_1, \dots, v_n$ are linearly independent because they forms a basis. Consequently $\alpha_1 - \beta_1 = \alpha_2 - \beta_2 = \cdots \alpha_n - \beta_n = 0$, i.e.\ $(\alpha_1, \dots, \alpha_n) = (\beta_1, \dots, \beta_n)$.
\end{proof}


\section{Proof of Theorem \ref{th:lin_dep}}\label{sec:proof_lin_dep}
	Notice that it suffices to prove the theorem for $k = n+1$ because if $x_1, \dots, x_{n+1}$ are linearly dependent, so are $x_1, \dots, x_{n+1}, \dots x_k$. We will therefore show for all $n \geq 1$
	\begin{align*}
		\mathcal{H}_n: \text{<< } &\text{For all } v_1, \dots, v_{n} \in V \ \text{and all } x_{1}, \dots x_{n+1} \in \Span(v_1, \dots, v_n), \\ &\text{the vectors } x_1, \dots, x_{n+1} \ \text{are linearly dependent. >>}
	\end{align*}
	\textbf{Base case:} $\mathcal{H}_1$ is true. Indeed, if $x_1,x_2 \in \Span(v_1)$, then there exists $\alpha_1, \alpha_2 \in \R$ such that $x_1 = \alpha_1 v_1$ and $x_2 = \alpha_2 v_1$. If $\alpha_1 = 0$ then $x_1 = 0$ and $x_1,x_2$ are therefore linearly dependent. 
	Otherwise if $\alpha_1 \neq 0$ then $v_1 = \frac{1}{\alpha_1} x_1$ which then gives $x_2 = \frac{\alpha_2}{\alpha_1} x_1$: $x_1,x_2$ are linearly dependent. 
	\\

	\noindent\textbf{Induction step:} We assume now that $\mathcal{H}_{n-1}$ holds for some $n \geq 2$ and we will show that $\mathcal{H}_{n}$ holds. We consider therefore $x_1, \dots, x_{n+1} \in \Span(v_1, \dots, v_n)$. We can find real numbers $\alpha_{i,j}$ such that
	$$
	\begin{array}{ccccccc}
		x_1 &=& \alpha_{1,1} v_1 &+& \cdots &+& \alpha_{1,n} v_n\\
		x_2 &=& \alpha_{2,1} v_1 &+& \cdots &+& \alpha_{2,n} v_n \\
		\vdots &&&&&& \\
		x_{n+1} &=& \alpha_{n+1,1} v_1 &+& \cdots &+& \alpha_{n+1,n} v_n.
	\end{array}
	$$
	We have to show that $x_1, \dots, x_{n+1}$ are linearly dependent.
	Let us consider the first line. If $\alpha_{1,1} = \alpha_{1,2} = \cdots = \alpha_{1,n} = 0$, then $x_1 = 0$ which gives then that $x_1, \dots, x_{n+1}$ are linearly dependent.
	Otherwise, there exists $j$ such that $\alpha_{1,j} \neq 0$. Without loss of generality we can assume that $\alpha_{1,1} \neq 0$. 
	$$
	\begin{array}{ccccccc}
		x_1 &=& \alpha_{1,1} v_1 &+& \cdots &+& \alpha_{1,n} v_n\\
		x_2 - \frac{\alpha_{2,1}}{\alpha_{1,1}} x_1 &=& 0 &+& \cdots &+& \alpha_{2,n} v_n - \frac{\alpha_{2,1}}{\alpha_{1,1}} \alpha_{1,n}v_n \\
		\vdots &&&&&& \\
		x_{n+1} - \frac{\alpha_{n+1,1}}{\alpha_{1,1}} x_1&=& 0 &+& \cdots &+& \alpha_{n+1,n} v_n- \frac{\alpha_{n+1,1}}{\alpha_{1,1}} \alpha_{1,n}v_n.
	\end{array}
	$$
	If we define $y_i \defeq x_i - \frac{\alpha_{i,1}}{\alpha_{1,1}} x_1$ for $i = 2, \dots, n+1$ we obtain have $y_i \in \Span(v_2, \dots, v_{n})$. We can now apply the induction hypothesis $\mathcal{H}_{n-1}$ to get that $y_2, \dots, y_{n+1}$ are linearly dependent. This means that there exists $\beta_2, \dots \beta_{n+1}$ that are not all zero, such that $\beta_2 y_2 + \cdots + \beta_{n+1} y_{n+1} = 0$ which finally gives
	$$
	\Big(- \beta_2 \frac{\alpha_{2,1}}{\alpha_1,1} - \cdots - \beta_{n+1} \frac{\alpha_{n+1,1}}{\alpha_{1,1}}\Big) x_1 + \beta_2 x_2 + \cdots + \beta_{n+1} x_{n+1} = 0.
	$$
	Since $\beta_2, \dots, \beta_{n+1}$ are not all zero we get that $x_1, \dots, x_{n+1}$ are linearly dependent. $\mathcal{H}_{n}$ is proved.

	\vspace{1cm}
	\centerline{\pgfornament[width=7cm]{71}}

%\bibliographystyle{plain}
%\bibliography{./references.bib}
\end{document}
