\documentclass{beamer}

\usepackage{../../latex_style/beamerthemeExecushares}
\usepackage{../../latex_style/notations}
\usepackage{dsfont}

\title{Recitation 10}
\author{Carles Domingo}
\date{Fall 2020}


\begin{document}

\frame{\titlepage} 

\setcounter{showProgressBar}{0}
\setcounter{showSlideNumbers}{1}

\begin{frame}[t]
\frametitle{Gradient and Hessian}
\vspace{-10pt}
\begin{definition}[Gradient and Hessian of a function]
Let $f : \R^n \rightarrow \R$ be a function. If it exists, its gradient at a point $x \in \R^n$ is defined as
\begin{equation*} 
\nabla f(x) = 
\begin{bmatrix}
\frac{\partial f}{\partial x_1}(x) \\
\vdots \\
\frac{\partial f}{\partial x_n}(x)
\end{bmatrix}
\end{equation*}
If it exists, its Hessian at a point $x \in \R^n$ is defined as
\begin{equation*} 
H f(x) = 
\begin{bmatrix}
\frac{\partial^2 f}{\partial x_1^2}(x) & \frac{\partial^2 f}{\partial x_1 \partial x_2}(x) & \cdots & \frac{\partial^2 f}{\partial x_1 \partial x_n}(x) \\ 
\frac{\partial^2 f}{\partial x_2 \partial x_1}(x) & \frac{\partial^2 f}{\partial x_2^2}(x) & \cdots & \frac{\partial^2 f}{\partial x_2 \partial x_n}(x) \\ 
\vdots &  \vdots & \cdots & \vdots \\ 
\frac{\partial^2 f}{\partial x_2 \partial x_1}(x) & \frac{\partial^2 f}{\partial x_2^2}(x) & \cdots & \frac{\partial^2 f}{\partial x_2 \partial x_n}(x)
\end{bmatrix}
\end{equation*}
\end{definition}
\end{frame}

\begin{frame}[t]
\frametitle{Convex sets and convex functions}
\vspace{-10pt}
\begin{definition}[Convex set]
A set $C \subseteq \R^n$ if for all $x,y \in C$, and all $\alpha \in [0,1]$,
\begin{equation*} 
\alpha x + (1-\alpha) y \in C.
\end{equation*}
\end{definition}

\begin{definition}[Convex function (and strictly convex function)]
A function $f : \R^n \rightarrow \R$ is convex if and only if for all $x, y \in \R^n$ and all $\alpha \in [0,1]$ it holds that 
\begin{align}
f(\alpha x + (1-\alpha) y) \leq \alpha f(x) + (1-\alpha) f(y).
\end{align}
It is strictly convex if moreover $\forall \alpha \in (0,1)$,
\begin{align}
f(\alpha x + (1-\alpha) y) < \alpha f(x) + (1-\alpha) f(y).
\end{align}
\end{definition}
\end{frame}

\begin{frame}[t] 
\frametitle{Convex sets}
\vspace{-10pt}
1. Which of the following sets are convex?
\begin{enumerate}
\item $\{x \in \R^2 : \|x\|=1\}$ 
\item $\{x \in \R^2 : \|x\| \leq 1\}$ 
\item $\{x \in \R^2 : \|x\| \geq 1\}$ 
\item $\{x \in \R^2 : \|x\| < 1\}$ 
\item $\{x \in \R^2 : v^\top x \geq a\}$  for fixed $v \in \R^2$ and $a \in \R$.  
\item $\{x \in \R^2 : v^\top x = a\}$  for fixed $v \in \R^2$ and $a \in \R$.  
\item $\{x \in \R^2 : x_2 \geq x_{1}^2 \}$ 
\item $\{x \in \R^2 : x_2 \leq x_{1}^2 \}$
\end{enumerate}
\pause
\pause
\pause
\end{frame}

\begin{frame}[t] 
\frametitle{Convex sets}
\vspace{-10pt}
1. Which of the following sets are convex?
\begin{enumerate}
\item $\{x \in \R^2 : \|x\|=1\}$ 
\item $\{x \in \R^2 : \|x\| \leq 1\}$ 
\item $\{x \in \R^2 : \|x\| \geq 1\}$ 
\item $\{x \in \R^2 : \|x\| < 1\}$ 
\item $\{x \in \R^2 : v^\top x \geq a\}$  for fixed $v \in \R^2$ and $a \in \R$.  
\item $\{x \in \R^2 : v^\top x = a\}$  for fixed $v \in \R^2$ and $a \in \R$.  
\item $\{x \in \R^2 : x_2 \geq x_{21} \}$ 
\item $\{x \in \R^2 : x_2 \leq x_{21} \}$
\end{enumerate}
\pause
\pause
\pause
\end{frame}

\begin{frame}[t] 
\frametitle{Convex functions and convex sets}
\vspace{-5pt}
For $f : \R^n \rightarrow \R$, define the epigraph $\text{epi}(f) \subset \R^{n+1}$ to be the set of all points above the graph of $f$: $\text{epi}(f) := \{(x,t) \in \R^{n+1} \ | \ t \geq f(x) \}.$
\begin{enumerate}
\item Prove that $f$ is convex if and only if $\text{epi}(f)$ is convex.
\item Prove that if $f,g$ are convex functions, then $h(x) =\max(f(x),g(x))$ is convex.
\end{enumerate}
\pause
\pause
\pause
\end{frame}

\begin{frame}[t]
\frametitle{Convex functions and Hessians}
\vspace{-5pt}
Reminder: If $f$ is a twice-differentiable function from $\R^n$ to $\R$, $f$ is convex if and only if its Hessian matrix $H_f(x)$ is positive semidefinite at all points $x \in \R^n$.

Reminder: If for all $x \in \R^n$, the Hessian matrix $H_f(x)$ is positive definite, then $f$ is strictly convex. Show that the reverse is not true, i.e. find a strictly convex twice-differentiable function such that the Hessian matrix $H_f(x)$ is not positive definite everywhere.
\pause
\pause
\end{frame}

\begin{frame}[t]
\frametitle{Convex functions and minima}
Let $f : \R^n \rightarrow \R$ be a convex function. Show that if for a certain $x \in \R^n$, there exists $\epsilon > 0$ such that $f(x) = \min \{f(y) \ | \ \|x-y\| < \epsilon \}$, then $f(x) = \min_{y \in \R^n} f(y)$.
\pause
\end{frame}

\begin{frame}[t]
\frametitle{Strictly convex functions \& minima}
Let $f : \R^n \rightarrow \R$ be a convex function. Show that there exists a unique $x^{\star}$ such that $f(x^{\star}) = \min_{y \in \R^n} f(y)$.
\pause
\end{frame}

\begin{frame}[t]
\frametitle{Gradients and Hessians}
\vspace{-10pt}
Calculate the gradients and the Hessians of the following functions $f : \R^n \rightarrow \R$:
\begin{enumerate}
\item $f(x) = \|x\|^2$.
\item $f(x) = \|Ax\|^2$.
\item $f(x) = x^\top A x$.
\end{enumerate}
For the functions 2 and 3, give necessary and sufficient conditions to have convexity and strict convexity.
\pause
\pause
\pause
\end{frame}

\begin{frame}[t]
\pause
\pause
\pause
\end{frame}


\end{document}


